Adaptive Sequential Model Selection

### 2015-05-10 paper.tex
tinkering on paper

We will consider two related problems. First, we will consider the problem of obtaining selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{0,k}:\; F\in M_{k-1}\quad \text{ vs. } \quad 
H_{1,k}:\; F\in M_\infty\setminus M_{k-1},
\]
while adjusting for the fact that the models are chosen adaptively. As we will see, {\em selected-model} tests can be dramatically more powerful than {\em saturated-model} tests at early steps in the model path when most of the signal variables have not yet entered the model.

The second half of the paper concerns {\em stopping rules} $\hk$, estimators of $k_0$. We will consider several stopping rules that operate on the full sequence $p_{[d]}$, including two powerful stopping rules proposed in \citet{gsell2013sequential}. These latter require guarantees on the joint law of the $p$-values. We will prove sufficient conditions for when the $p$-values are Gaussian.

Although we will focus most of our attention and examples on the case of selecting a set of predictors in linear regression, essentially all of our results apply in generic exponential family models.

@@

There is some ambiguity involved in deciding how to generalize $z$- and $t$-tests to the selective case. For example, \citet{gsell2013sequential} describe three different null hypotheses that we could consider testing at step $k$. As discussed at some length in \citet{fithian2014optimal}, there are several different models that

@@

\section{Sequential Inference}

@@

We will consider stopping rules that depend on the sequence $p_1,\ldots,p_d$ of $p$-values, which have been studied in the literature by \WFcomment{literature review here.} We will focus on three such stopping rules: simple stop, proposed by \WFcomment{whom?}, and the more refined strong stop and forward stop, proposed by \citet{gsell2013sequential}.

While strong stop and forward stop tend to give more powerful stopping rules, they require independence among the $p$-values. Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which selective $p$-values are mutually independent. We will see that, typically, selected-model $p$-values are sequentially independent while saturated-model $p$-values are not.

@@

The most obvious sequential procedure is to reject at each step until the first time $p_k > \alpha$, which we can formalize as
\[
\hk_B(Y) = \min\left\{k \in \{1,\ldots,m\} :\;
  p_k > \alpha\right\} - 1
\]
We will call this procedure {\em basic stop}.

@@

\begin{model}[Generalized Sparse Model]\label{mod:genSparse}
  Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^p$:
  \[
  M_\infty = \{F_\theta:\; \theta \in \Theta\}.
  \]
  For subsets $E\sub [p]$ define the submodel with active coefficients $E$ as follows:
  \[
  \Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
  \quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
  \]
  Let $\ell(\theta; Y)$ denote the log-likelihood for model
  $M_\infty$, and define the restricted log-likelihood
  \[
  \ell_E(\theta; Y) = \left\{\begin{matrix} \ell(\theta; \;Y) 
      & \theta \in \Theta(E)\\ -\infty  & 
      \mathrm{ otherwise.}\end{matrix}\right. 
  \]
\end{model}
Although we will make no formal assumptions about the form of $M_\infty$, the results of this section are only useful in cases where we can actually construct single-step $p$-values. At present, that caveat  largely restricts us to using exponential family models.

>>>>

We will consider two related problems. First, we will consider the problem of obtaining selective single-step $p$-values, where $p_k$ is a $p$-value for testing
\[
H_{0,k}:\; F\in M_{k-1}\quad \text{ vs. } \quad 
H_{1,k}:\; F\in M_\infty\setminus M_{k-1},
\]
while adjusting for the fact that the models are chosen adaptively. In most of our development we will make few formal assumptions about the form of the candidate models. Still, our results are only useful in cases where we can actually construct single-step $p$-values. At present, that means our applications are largely restricted to exponential family models.

Most of our examples will focus on the important and familiar problem of adaptively selecting a set of predictors in linear regression models. As we will see, {\em selected-model} tests can be dramatically more powerful than {\em saturated-model} tests at early steps in the model path when most of the signal variables have not yet entered the model.

Section~\ref{sec:sequential} concerns {\em stopping rules} $\hk$, estimators of $k_0$. We will consider several stopping rules that operate on the full sequence $p_{[d]}$, including two powerful stopping rules proposed in \citet{gsell2013sequential}. These latter require guarantees on the joint law of the $p$-values. We will prove sufficient conditions for when the $p$-values are Gaussian.

@@

There is some ambiguity involved in deciding how to generalize $z$- and $t$-tests to the selective case. For example, \citet{gsell2013sequential} describe three different null hypotheses that we could consider testing at step $k$ in the case of linear regression under the upper model $M_\infty:\; Y\sim \cN(\mu, \sigma^2 I)$:
\begin{description}
\item[Complete Null:] The previous model, $M_{k-1}$, is correct; i.e., that $\mu = X_{E_{k-1}}\beta$ for some $\beta$.
\item[Incremental Null:] $M_{k-1}$ may not be correct, but $M_k$ is no improvement: i.e., that $X_EX_E\mu = \proj_{X_{E_{k-1}}}\mu$
\end{description}
. \citet{fithian2014optimal} discusses several different tests

@@

\section{Sequential Inference}\label{sec:sequential}

@@

We will consider stopping rules that depend on the sequence $p_1,\ldots,p_d$ of $p$-values, which have been studied in the literature by \WFcomment{literature review here.} We will focus on three such stopping rules: Basic Stop, proposed by \WFcomment{whom?}, and the more refined Strong Stop and Forward Stop, proposed by \citet{gsell2013sequential}.

While Strong Stop and Forward Stop tend to give more powerful stopping rules, they require independence among the $p$-values. Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which selective $p$-values are mutually independent. We will see that, typically, selected-model $p$-values are sequentially independent while saturated-model $p$-values are not.

@@

The most obvious sequential procedure is to reject at each step until the first time $p_k > \alpha$, which we can formalize as
\[
\hk_B(Y) = \min\left\{k \in \{1,\ldots,m\} :\;
  p_k > \alpha\right\} - 1
\]
We will call this procedure {\em Basic Stop}.

@@

\begin{model}[Generalized Sparse Model]\label{mod:genSparse}
  Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^p$:
  \[
  M_\infty = \{F_\theta:\; \theta \in \Theta\}.
  \]
  For subsets $E\sub [p]$ define the submodel with active coefficients $E$ as follows:
  \[
  \Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
  \quad M(E) = \{F_\theta:\; \theta\in \Theta_E\}.
  \]
  Let $\ell(\theta; Y)$ denote the log-likelihood for model
  $M_\infty$, and define the restricted log-likelihood
  \[
  \ell_E(\theta; Y) = \left\{\begin{matrix} \ell(\theta; \;Y) 
      & \theta \in \Theta(E)\\ -\infty  & 
      \mathrm{ otherwise.}\end{matrix}\right. 
  \]
\end{model}
